---
title: "PA1_template.Rmd"
author: "Sarai Nwagbaraocha"
date: "April 18, 2016"
output: html_document
---

###About
This was the first project for the **Reproducible Research** course in Coursera's Data Science specialization track. The purpose of the project was to answer a series of questions using data collected from a [FitBit](http://en.wikipedia.org/wiki/Fitbit).


##Synopsis
The purpose of this project was to practice:

* loading and preprocessing data
* imputing missing values
* interpreting data to answer research questions

## Data
The data for this assignment was downloaded from the course web
site:

* Dataset: [Activity monitoring data](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]

The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing
    values are coded as `NA`)

* **date**: The date on which the measurement was taken in YYYY-MM-DD
    format

* **interval**: Identifier for the 5-minute interval in which
    measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data

Download, unzip and load data into data frame `data`.

```{r , echo=FALSE}
data <- read.csv("activity.csv", header = TRUE, sep = ',', colClasses = c("numeric", "character",
                                                                          "integer"))
```

## What is mean total number of steps taken per day?
Sum steps by day, create Histogram, and calculate mean and median.

```{r steps, echo=TRUE}
steps_by_day <- aggregate(steps ~ date, data, sum)
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")
rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)
```

## What is the average daily activity pattern?

* Calculate average steps for each interval for all days. 
* Plot the Average Number Steps per Day by Interval. 
* Find interval with most average steps.

```{r avg daily activity, echo=TRUE}
steps_by_interval <- aggregate(steps ~ interval, data, mean)

plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")

max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
```

## Impute missing values. Compare imputed to non-imputed data.
Missing data needed to be imputed. Only a simple imputation approach was required for this assignment. 
Missing values were imputed by inserting the average for each interval. Thus, if interval 10 was missing on 10-02-2012, the average for that interval for all days (0.1320755), replaced the NA. 

```{r missing values, echo=TRUE}
incomplete <- sum(!complete.cases(data))
imputed_data <- transform(data, steps = ifelse(is.na(data$steps), steps_by_interval$steps[match(data$interval, steps_by_interval$interval)], data$steps))

imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0

steps_by_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")

#Create Histogram to show difference. 
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="red", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)

rmean.i <- mean(steps_by_day_i$steps)
rmedian.i <- median(steps_by_day_i$steps)

mean_diff <- rmean.i - rmean
med_diff <- rmedian.i - rmedian

total_diff <- sum(steps_by_day_i$steps) - sum(steps_by_day$steps)
```

The impact of imputing missing data with the average number of steps in the same 5-min interval is that both the mean and the median are equal to the same value: 10766.

```{r wkdays wkends, echo=TRUE}

weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))

steps_by_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)

library(lattice)

xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")

```

